Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-08-20 paint-contour-tree.tex
Structural changes to section 5

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
\newcommand{\pth}[2][\!]{#1\left({#2}\right)}

@@

\pagebreak
\section{Lower Bound}

@@

\subsection{Path Decomposition Background}
\label{sec:background}
We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
running time of our algorithm.

\begin{definition}
 Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
 is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
 For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.

@@

% \begin{observation}
% DELETE IF NOT NEEDED:
% Note that $f_{path}(P(T)) = \Theta( \sum_{v\in V} l_v )$, where $l_v$ is the distance from $v$ to the leaf of its containing path in $P(T)$.  
% That is, conceptually we can think of each vertex as charging the $\log$ of the length to its path's leaf.
% \end{observation}

@@

\newcommand{\pathTree}{P_{\mathcal{T}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
\begin{definition}
Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

We say two path in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree. 
\end{definition}

%\begin{definition}
%We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
%Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
%$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
%Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
%\end{definition}

@@

 Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
\end{lemma}
\begin{proof}
 %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
 %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
 %Observe that $\sum_{i=0}^k d_i = l_p$.
 %
 % 
 %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

@@

Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.

\Ben{There is a fundamental flaw in the following lemma and corollary.}

\begin{lemma}
Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
$\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
\end{lemma}
\begin{proof}
By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
 As $x\log x$ is a convex function, Jensen's inequality then implies 
 $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
\end{proof}

\begin{corollary}
\label{cor:heavyLight}
 Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
 Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
\end{corollary}
\begin{proof}
 Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
 $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
 Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
 Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.

\end{proof}

\subsection{Lower Bound by Path Decomposition}
We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.

@@

\subsection{Upper Bound by Path Decomposition}
\Ben{This section needs revision.}

From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a particular path decomposition.

Specifically, for a given valid coloring $\chi$ of a tree $T$, we describe a path decomposition for which 
we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about path decompositions from 
\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
 Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
 contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
 of all light reachable paths from $p$ the \emph{light path tree} of $p$.
\end{definition}

Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
We now define a weight function $w(v)$ for all $v\in \mathcal{L}$.  Such that $w(v)=\Theta(|H_v|)$.  
For this alternative quantity we will be able to show BLAH BLAH BLAH.

%Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely 
%the total sum of the sizes of the heaps at the roots of all paths in the light path tree is on the 
%order of the size of the heap at the root of $r$.  ADD THE ARGUMENT saying how this fact will be used.

@@

\begin{lemma}
\label{lem:distribution}
Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/20$
\end{lemma}

@@

By assumption from the lemma statement we have $w(u_{i+1})\leq (3/5) w(u_i)$.  Therefore by the above calculation, 
the total reassigned to $u_m$ is

\[
 \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{3}}^iw(u_m)} 
 \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{3}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{20}
\]

@@

\[
|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/20 \geq \frac{4}{5} w(v)
\]
\end{claim}

\begin{lemma}
Let $\mathcal{L}$ be the light path tree of some path $r$.  
Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
\end{lemma}
\begin{proof}
We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ for any two paths 
$x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.

Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
(i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{geometric},

\end{proof}

>>>>

\newcommand{\XSays}[2]{{
      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
            #1:} }
      \textcolor{red}{#2}
      \marginpar{\textcolor{blue}{#1}}
      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
      }
   }
\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
\newcommand{\remove}[1]{}
\newcommand{\pth}[2][\!]{#1\left({#2}\right)}

@@

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{T}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
\section{Lower Bound}

@@

To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
we then relate path decompositions costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

\begin{definition}
\label{def:path}
 Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
 is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
 For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.

@@

Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
the path decomposition $P(T)$, and denote it by $\pathTree(T)$.

We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree.

\remove{
\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}

@@

@@

 Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
\end{lemma}
\begin{proof} 
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

@@

% \Ben{There is a fundamental flaw in the following lemma and corollary.  
% As such, the remainder of this subsection will likely be deleted.}
% 
% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
% 
% \begin{lemma}
% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
% \end{lemma}
% \begin{proof}
% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
%  As $x\log x$ is a convex function, Jensen's inequality then implies 
%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
% \end{proof}
% 
% 
% \begin{corollary}
% \label{cor:heavyLight}
%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
% \end{corollary}
% \begin{proof}
%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
%  
% \end{proof}

}

\subsection{Lower Bound by Path Decomposition}

In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.

We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.

@@

\subsection{Upper Bound by Path Decomposition}
\Ben{This section is a work in progress.}

From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
which we will call the maximum path decomposition (defined below).
Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
%using the observations about path decompositions from 
%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.

\subsubsection{Heavy and light paths}

Before describing the maximum path decomposition, we introduce the notion of heavy and light paths 
(which will help motivate what the decomposition should be).

Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
we will divide paths into two categories, called heavy and light, which we show below in \Lem{pathBounds} roughly corresponds 
to whether a path can pay for itself or not.

The following lemma demonstrates that heavy paths can ``pay'' for themselves.

We are now ready to define the maximum path decomposition, which will allow us to recharge the light paths to heavy paths.

\subsubsection{The decomposition and its main property}

@@

\begin{definition}
 Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
 Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
 contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
 of all light reachable paths from $p$ the \emph{light path tree} of $p$, and denote is $\mathcal{L}(p)$.
\end{definition}

In the following subsection we show how to exploit this geometric series behavior in light path trees.

\subsubsection{An asymptotically equivalent weight function}
Let $\mathcal{L} = \mathcal{L}(r)$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
The intuitive idea is that due to the geometric decreasing behavior of light path trees described in the previous subsection, 
one can show that for the entire light path tree our algorithm pays roughly the cost due to the root path $r$.  This argument 
is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.

@@

\begin{lemma}
\label{lem:distribution}
Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
\end{lemma}

@@

By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
the total reassigned to $u_m$ is

\[
 \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
 \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
\]

@@

\[
|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
\]
\end{claim}

\begin{lemma}
Let $\mathcal{L}$ be the light path tree of some path $r$.  
Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
\end{lemma}
% \begin{proof}
% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
% of $v$ in $\mathcal{L}$.
% 
% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
% Combining this fact with \Lem{geometric} gives,
% \[
% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
% \]
% 
% \end{proof}
%
\begin{proof}
By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
\[
w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
\]
However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
fractional decrease always holds implying the claim always holds.
\end{proof}

\begin{corollary}
 Let $\mathcal{L}$ be a light path tree of a light path $r$.  Then for any $v\in \mathcal{L}$, $w(v) = O(|H_v|)$.
\end{corollary}
